---
title: "BIO 356 / BEE 587 Lab Number - Lab Name"
author: "Brian Wang"
date: "4/15/23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)

# don't forget to copy/paste here all of the libraries required for this week's lab
library(tidyverse)
library(gridExtra) ## for plotting function grid.arrange()
library(deSolve) ## for predator-prey-resource ODE model
library(magrittr) ## for pipe symbol %<>%
library(knitr) ## for table-viewing function kable()

theme_set(theme_bw())
theme_update(
  panel.grid = element_blank(),
  aspect.ratio = 1
)
```

## Q1
Since there are no consumers, the equilibrium population of the resource would be the resource carrying capacity K.

## Q2
I don't know.

## Q3
r1n1(1 - N1/k1) since N2 would be 0. This is an logistic curve.

## Q4
consumer 1 would experience exponential growth. r1 would be the relative rate of growth.

```{r}
LV_Model = 
  function(
    initial_N,
    intrinsic_growth_rate,
    carrying_capacity,
    competition_coefficient,
    final_time,
    time_step
  ){
      LV = 
        function(t, state, parameters){
           with(as.list(c(state, parameters)), {
             N = c(N1, N2)
             dNdt = as.numeric(r * N * (1 - A %*% N / K))
             list(c(dNdt[1], dNdt[2]))
           })
          }
      
      times = seq(0, final_time, by = time_step)
      
      parameters = 
        list(
          r = intrinsic_growth_rate,
          K = carrying_capacity,
          A = 
            matrix(
              c(1, competition_coefficient[1], competition_coefficient[2], 1), 
              byrow = TRUE, 
              nrow = 2
            )
        )
      
      state = c(N1 = initial_N[1], N2 = initial_N[2])
      
      out = ode(y = state, times = times, func = LV, parms = parameters)
      
      return(
        list(
          parameters = 
            list(
              intrinsic_growth_rate, 
              carrying_capacity, 
              competition_coefficient
            ),
          initial_conditions = initial_N,
          state = out
        )
      )
   }

Plot_LV = 
  function(model){
    as.data.frame(model$state) |>
    pivot_longer(-time, names_to = 'species', values_to = 'N') |>
    ggplot(aes(time, N, group = species, color = species)) +
    geom_line(size = 1) +
    expand_limits(y = 0) +
    labs(
      x = 'Time',
      y = 'Abundance'
    )
  }
```

```{r}
model = 
  LV_Model(
    initial_N = c(30, 30), 
    intrinsic_growth_rate = c(1, 1), 
    carrying_capacity = c(50, 50), 
    competition_coefficient = c(1.01, 1/1.01), 
    final_time = 1000, 
    time_step = 1
  )

# plot results
Plot_LV(model)
```
## Q5
Both species did not coexist in the end. N1 went extinct because N2 had a better competition coefficient than N1.

```{r}
model = 
  LV_Model(
    initial_N = c(30, 30), 
    intrinsic_growth_rate = c(1, 1), 
    carrying_capacity = c(50, 50), 
    competition_coefficient = c(1/1.01, 1.01), 
    final_time = 1000, 
    time_step = 1
  )

# plot results
Plot_LV(model)
```
## Q6
swaping the competition coefficient will result in N1 outcompeting N2.

```{r}
model = 
  LV_Model(
    initial_N = c(30, 30), 
    intrinsic_growth_rate = c(1, 1), 
    carrying_capacity = c(50, 50), 
    competition_coefficient = c(0.3, 0.6), 
    final_time = 1000, 
    time_step = 1
  )

# plot results
Plot_LV(model)
```
## Q7
Both species were able to coexist under these conditions.

## Q8
a) Species 2 regulates itself more than species 1. 
b) species 1 regulates species 2 more than species 2 itself.

```{r}
initial_N1 = 7
initial_N2 = 5
r1 = 1
r2 = 1
K2 = 30
alpha_12 = 1.1
alpha_21 = 1/alpha_12
K1 = alpha_12 * K2

# call the Lotka-Volterra model 
model = 
  LV_Model(
    initial_N = c(initial_N1, initial_N2), 
    intrinsic_growth_rate = c(r1, r2), 
    carrying_capacity = c(K1, K2), 
    competition_coefficient = c(alpha_12, alpha_21), 
    final_time = 50, 
    time_step = .1
  )

N1_final = model$state[nrow(model$state), 2]
N2_final = model$state[nrow(model$state), 3]

# plot results
Plot_LV(model) +
  ggtitle(paste('Final abundances: N1 =', round(N1_final), '     N2 =', round(N2_final)))
```
## Q9
Final abundances were not the same as the other initial abundances.

## Q10
two different species of bird both rely on seeds and bugs as resources.

## Q11
the preferred resource of consumer 1 is the first resourse while the prefferred resource of consumer 2 is the second resource.

## Q12
the scenario where each consumer prefers a different resource because then it would allow for less overlap

## Q13
Scenario 1 is intraspecific competition while scenario 2 is interspecific

```{r}
## set parameter values
r = 0.1
K1 = 100
K2 = 100
alpha12 = 0.75
alpha21 = 0.75

## calculate predicted equilibrium abundances
N1_equil = (K1 - alpha12 * K2) / (1 - alpha12 * alpha21)
N2_equil = (K2 - alpha21 * K1) / (1 - alpha12 * alpha21)

## call the Lotka-Volterra model 
model = 
  LV_Model(
    initial_N = c(100, 100), 
    intrinsic_growth_rate = c(r, r), 
    carrying_capacity = c(K1, K2), 
    competition_coefficient = c(alpha12, alpha21), 
    final_time = 1000, 
    time_step = 1
  )

## plot results, adding horizontal dashed lines marking the predicted equilibrium abundances
Plot_LV(model) +
  ggtitle(paste('Predicted equilibrium abundances: N1 =', round(N1_equil), '     N2 =', round(N2_equil)))

model = 
  LV_Model(
    initial_N = c(100, 90), 
    intrinsic_growth_rate = c(r, r), 
    carrying_capacity = c(K1, K2), 
    competition_coefficient = c(alpha12, alpha21), 
    final_time = 1000, 
    time_step = 1
  )

## plot results, adding horizontal dashed lines marking the predicted equilibrium abundances
Plot_LV(model) +
  ggtitle(paste('Predicted equilibrium abundances: N1 =', round(N1_equil), '     N2 =', round(N2_equil)))


## set parameter values
r = 0.1
K1 = 100
K2 = 100
alpha12 = 1.25
alpha21 = 1.25

## calculate predicted equilibrium abundances
N1_equil = (K1 - alpha12 * K2) / (1 - alpha12 * alpha21)
N2_equil = (K2 - alpha21 * K1) / (1 - alpha12 * alpha21)

## call the Lotka-Volterra model 
model = 
  LV_Model(
    initial_N = c(100, 100), 
    intrinsic_growth_rate = c(r, r), 
    carrying_capacity = c(K1, K2), 
    competition_coefficient = c(alpha12, alpha21), 
    final_time = 1000, 
    time_step = 1
  )

## plot results, adding horizontal dashed lines marking the predicted equilibrium abundances
Plot_LV(model) +
  ggtitle(paste('Predicted equilibrium abundances: N1 =', round(N1_equil), '     N2 =', round(N2_equil)))

## call the Lotka-Volterra model 
model = 
  LV_Model(
    initial_N = c(100, 90), 
    intrinsic_growth_rate = c(r, r), 
    carrying_capacity = c(K1, K2), 
    competition_coefficient = c(alpha12, alpha21), 
    final_time = 1000, 
    time_step = 1
  )

## plot results, adding horizontal dashed lines marking the predicted equilibrium abundances
Plot_LV(model) +
  ggtitle(paste('Predicted equilibrium abundances: N1 =', round(N1_equil), '     N2 =', round(N2_equil)))
```
## Q14
The first two plots shows what happens when both relative abundances are less than 1, resulting in coexistence. The second two plot shows what happens when both relative abundances are more than 1.

## Q15
scenario 2 and 4 has priority effects where species whose initial population is higher will be the one that out competes the other species while scenario 1 and 2 shows stable coexistence.

## Q16
species 1 consumes more resource 1 and 2.

## Q17
this means that species 2 will get out competed by species 1

## Q18
the competitiveness of species 2 is less than the competitiveness of species 1.

## Q19
these two species cannot reliably coexist.

## Q20
---?

## Q21
Pc will have outcompete PB.

