---
title: "BIO 356 / BEE 587 0 part 2 - Intro to R"
author: "Brian Wang"
date: "1/29/24"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)

# don't forget to copy/paste here all of the libraries required for this week's lab
library(tidyverse)
library(rio)

theme_set(theme_bw())
```




## Q14

Gives the average mpg of the vehicles in mtcars. The value outputted is 20.090

```{r}
mean(mtcars$mpg)
```

## Q15

Average is 3.1895

```{r}
var(mtcars$cyl)
```

## Q16

Code Below

```{r}
characters = 
  tibble(
    name = c('Calvin', 'Hobbes', 'Garfield', 'Popeye', 'Bugs Bunny', 'Lisa Simpson', 'Woodstock'),
    species = c('human', 'tiger', 'cat', 'human', 'rabbit', 'human', 'canary'),
    comic_strip = c(TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, TRUE),
    year_created = c(1985, 1985, 1978, 1929, 1940, 1987, 1967)
  )

characters |> arrange(year_created)
```

## Q17

Code below.

```{r}
mtcars |> 
  rownames_to_column(var = "model") |>
  as_tibble() |>
  select(model, mpg, cyl, hp) |>
  mutate(kml = mpg * 0.43) |>
  filter(mpg > 22)
```

## Q18

new_abundance is 16098 with a standard error of 834.

```{r}
github_address = 'https://github.com/rafaeldandrea/BIO-356/blob/master/gray_whales.xlsx?raw=true'

whale_dataset = 
  rio::import(github_address, range = 'B6:G28') |> 
  as_tibble()

whale_dataset
```

## Q19

We would get an error if we piped it in like this "plot_whales |> goem_line". instead use the "+" and include it when the plot is being created.

```{r}
plot_whales = 
  whale_dataset |> 
  ggplot(aes(End_dates, New_abundance)) + 
  geom_point() +
  geom_line()

plot_whales
```

## Q20
```{r}
plot_whales = 
  whale_dataset |> 
  ggplot(aes(End_dates, New_abundance)) + 
  geom_line(color = 'green') +
  geom_errorbar(
    aes(
      x = End_dates, 
      ymin = New_abundance + New_SE, 
      ymax = New_abundance - New_SE
    ),
    color = 'blue'
  ) +
  geom_point(color = 'red') +
  labs(
    x = 'Calendar year',
    y = 'Population size'
  ) +
  theme(aspect.ratio = 1) +
  ggtitle('Whale population estimates')

plot_whales
```

## Q21

it shows the result below. This is because '=' is an assignment operator while '==' checks for equivalence.

```{r error = TRUE}
if(2 > 3){
  print('Two is greater than three')
} else if(2 = 3){ 
  print('Two is equal to three')
} else {
  print('Two is less than three')
}
```

## Q22

```{r}
for (i in 1:5) {
 if (i %% 2 == 1) {
  print(i)
 }
}
```

## Q23

Changing the growth rate to a value below 1.0 will result in a decrease of population.

```{r}
lambda = 0.9
N = 50
N_vector = N
for(year in 1:20){
  N = round(lambda * N)
  N_vector = c(N_vector, N)
}

data = 
  tibble(
    year = 0:20,
    abundance = N_vector
  )

plot = 
  data |>
  ggplot(aes(year, abundance)) +
  geom_line(color = 'grey') +
  geom_point()

plot
```